---
title: "My Final Project Template"
author: Your name here
subtitle: Subtitle here if desired
date: today
date-format: long
---

# Introduction

\[\~ 200 words\]

Clearly stated background and questions / hypotheses / problems being addressed. Sets up the analysis in an interesting and compelling way. Include figures if you like.

# Materials and methods

\[\~ 200 words\]

Narrative: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.

You can do bullets like this:

-   The first most important thing
-   The second most important thing
-   The third most important thing

You can do numbers like this:

1.  The first most important thing
2.  The second most important thing
3.  The third most important thing

See <http://rmarkdown.rstudio.com/> for all the amazing things you can do.

Here's my first code chunk.

```{r}

```


Load any required packages in a code chunk (you may need to install some packages):

```{r, message=F, warning=F}

```



# Conclusions

\[\~200 words\]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
